Stage-Splitting a Modal Language

### 2014-10-20 paper/example.tex
working on example.tex

@@

But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
the staged version of quickselect in \ref{fig:qsstaged} is virtually identical to the unstaged version in \ref{fig:quickselect}.
The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
but that would not be particularly interesting since it would result in an effectively trivial split 
that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
This process is known as {\em binding time analysis}, and we do not consider it here.
Instead, we assume that all input programs are properly staged according to some programmer's intent.

>>>>

fun qSelect (l : list, k : int) : int = 
  case l of
    Empty => 0
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
        case compare k n of
          LT => qSelect (left, k)
        | EQ => h
        | GT => qSelect (right, k-n-1)
\end{lstlisting}
\caption{Unstaged implementation of quickselect.}
\label{fig:quickselect}
\label{fig:qs-unstaged}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\begin{lstlisting} 
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...

fun qsStaged (l : list, k : $`2`int`1`) : $`2`int`1` = 
  case l of
    Empty => next {`2`0`1`}
  | Cons (h,t) => 
      let (left,right,n) = partition h t in
      next{
        `2`let n = hold{`1`n`2`} in
          case compare prev{`1`k`2`} n of
            LT => prev {`1`qsStaged (left k)`2`}
          | EQ => hold {`1`h`2`}
          | GT => prev {`1`qsStaged (right, 
                             next{prev{k}-n-1)}`2`}`1`
      }`
\end{lstlisting}
\caption{Staged implementation of quickselect in \lang.}
\label{fig:qsstaged}
\label{fig:qs-staged}
\end{subfigure}
\caption{Quickselect: traditional and staged.}
\end{figure*}

Staging is a technique for using language constructs and types to
explicitly staged a computation according to the frequency or the
arrival time of data.  As an example, suppose that we wish to issue
order statistics queries on a collection of items represented as a
list \texttt{l}.  For this tasks, we can use the expected linear time,
quickselect algorithm~\cite{quickselect}, which takes a
list~\texttt{l} and a rank~\texttt{k} and returns the element with
rank~\texttt{k} in the list.  To this end, the algorithm, whose code
is shown in ~\figref{qs-unstaged} partitions the list into two by
using the first element as a pivot~\footnote{We assume that the list
  is prepermuted to guarantee the expected linear time behavior.} and
recurs down to one of the ``halves'' to find the desired element.  The
``half'' to be considered recursively next is determined by the
relationship of~\texttt{k} to the size of the first half~\texttt{n},
which is returned by the function~\texttt{partition}.

Continuing on our example, consider an application where we perform
many order statistics queries using the same collection~\texttt{l}, but
potentially with different ranks~\cite{k}.  To take advantage of this,
we can {\em stage} quickselect so that the parts of the computation
that depend on the collection~\texttt{l} is performed once, and the
desired result is computed by using~\texttt{k} and the results from
this prior stage.

Prior work proposed staging for exactly these kinds of applications
and proposed techniques for writing staged
programs~\cite{staging}. Using techniques essentially the same as
prior work, we can stage our quickselect code by using staged typing
and inserting three keywords~\texttt{next, prev, hold} into the
appropriate places in the code as shown in \figref{qs-staged}. To
write the staged version, the programmer first observes that the
rank~\texttt{k} has now the type $\fut\rmint$ (read ``circle int'')
rather than $\rmint$, indicating that it will be available in the
second stage.  All types that are not of circle type are available in
the first stage.  Naturally the output type, representing the $k$th
largest element of the list, is also $\fut\rmint$, since it is not
computed until the second stage. Having determined the type signature,
the programmer can then insert $\next$ and $\prev$ constructs into the
code. The construct $\next$ occurs in a stage \bbone\ context and
indicates that the contents of its block are stage \bbtwo, whereas
$\prev$ occurs in a stage \bbtwo\ context and indicates that the
contents of its block is stage \bbone.  (We adopt the convention that
the top-level context is stage \bbone.)  The output type of a $\next$
block is the $\fut$'d version of the type of its stage
\bbtwo\ contents.  For example, \verb|next{0}| from above has the type
$\fut\rmint$.  Correspondingly, $\prev$ requires that its stage
\bbone\ contents have a $\fut$ type, and it eliminates the wrapper.
For example, \verb|prev{k}| from above has type $\rmint$ at stage
\bbtwo, since $k$ is bound to a $\fut\rmint$ at stage
\bbone. Informally speaking, the
$\next$ construct enables operating on second stage computations in
the first stage essentially by treating them as boxes that can be
moved around, stored in data structures, and passed to functions, and
the $\prev$ construct enables unwrapping the contents of such a box to
use in the second stage.

\ur{$\pause$ should be called ``promote'' or something like that. I
  believe this is what Taha calls cross stage persistence.  THat work should
  be cited.}

For convenience, the code also uses $\pause$ blocks.  This construct
has same stage signature as $\prev$, but instead of ``unwrapping"
$\fut$ types it simply promotes integers from stage~\bbone\ to
stage~\bbtwo.
\footnote{It will turn out that $\pause$ is implementable---though it takes some effort---from our other language features.
We instead provide it as a primitive to shorten examples.  
Furthermore, it would be wise to extend $\pause$ to other base types, if we had them, and to products and sums thereof.
This is related to the notion of {\em mobility} in \cite{murphy05} and
{\em stability} in \cite{krishnaswami13}.}

We note that given the type signature of staged quickselect, there are
several ways to insert the staging annotations.  The version we
presented here maximizes the work done at stage~\bbone.

\ur{I don't like fuzzy words like hygenic that dont mean anything.
  What is the point of using them?  A paper ought to be written for
  the reader not as a note to self.}

%%%
%% There has been the extensive research into the question of how to
%% automatically add staging annotations to unstaged code.  This process
%% is known as {\em binding time analysis}, and we do not consider it
%% here.  Instead, we assume that all input programs are properly staged
%% according to some programmer's intent.

Prior work on meta programming allows us to evaluate such staged code
in a staged fashion by first providing the first-stage arguments and
generating a piece of code that is specialized for the arguments.  For
example, we can evaluate the staged quickselect by providing a
list~\texttt{l} to obtain a piece of code that then can be used with
different ranks (\texttt{k} values) in the second stage.  Such a
specialized code essentially performs all the first-stage computations
first and uses their results to improve the efficiency in the second
stage, where we issue many queries.  The type system guarantees that
all first-stage computations depend only on the first stage values and
can indeed be computed in the absence of second-stage values.

This staged evaluation technique corresponds to {\em partial
  evaluation} or {\em program specialization} and can bring out
dramatic improvements in the efficiency of programs.  It has however
one major limitation: since it requires the first stage arguments to
be available, it cannot be performed at compile time, except of course
when the first-stage values are available at compile time.  Since the
evaluation in the first stage produces a piece of code that must also
be compiled or interpreted to execute in the second stage, the
approach can lead to code blowup and tends to be relatively expensive
in practice.  Due to these reasons, the applications of partial
evaluation are limited.  For example, in our quickselect example, it
is unlikely to know the list~\texttt{l} that we will be operation on
at compile time. Partial evaluation dynamically at run time is
difficult because the first stage generates code that is linear in the
size of the list, whose generation, dynamic compilation or
interpretation can be excruciatingly slow.

Knoblock and Ruf~\cite{KR} proposed  the {\em data specialization} technique

limitations of

Prior work on

To illustrate this, we present colorized version of \texttt{qSelect}:

\begin{figure}

\begin{subfigure}{0.5\textwidth}
\begin{lstlisting}
1`datatype list = Empty | Cons of int * list
fun partition (p : int, l : list) = ...`

2`fun qSelect2 (p : tree, k : int) : int = 
  case unroll p of
    Leaf => 0
  | Branch (n,h,p1,p2) => 
      case compare k n of
        LT => qSelect2 (p1,k)
      | EQ => h
      | GT => qSelect2 (p2,k-n-1)`
\end{lstlisting}
\caption{Split (two-phase) implementation of quickselect.}
\label{fig:qssplit}
\end{subfigure}
\caption{Caption place holder}
\end{figure}

\begin{comment}

%% Umut: this is the earlier version.  I am reorganizing this whole
%% section.

@@

But for the $\fut$, $\next$, $\prev$, and $\pause$ constructs, 
the staged version of quickselect in \ref{fig:qsstaged} is virtually identical to the unstaged version in \ref{fig:quickselect}.
The constructs that were added were placed in order to maximize the work done at stage~\bbone\ while still conforming to the type signature.
It would have also been valid to simply move the whole input list unchanged into stage~\bbtwo\ at the very beginning, 
but that would not be particularly interesting since it would result in an effectively trivial split 
that's just the identity at stage~\bbone\ and plus quickselect at stage~\bbtwo.
There has been the extensive research into the question of how to automatically add staging annotations to unstaged code.
This process is known as {\em binding time analysis}, and we do not consider it here.
Instead, we assume that all input programs are properly staged according to some programmer's intent.
\end{comment}
